---
title: "Supplementary material 3" 
subtitle: "The Big Picture: a Comparative Analysis of the Social Determinants of Health Across Europe"
output: pdf_document
date: ""
editor_options: 
  chunk_output_type: console
---

```{r}
# Libraries loading
library(haven)
library(dplyr)
library(ggplot2)
library(tidyr)
library(lme4)
library(readxl)
library(labelled)
```

# Data arrangements

## Loading of the original dataset. 

```{r}
# Data downloaded from: https://doi.org/10.21338/ess11e04_1
ESS11FullRaw <- read.csv(file.path("datasets_raw","ESS11e04_1.csv"))
```

## Recoding of some of the variables of analysis

```{r}
# NAs for some variables
ESS11FullRaw$cgtsmok[ESS11FullRaw$cgtsmok == 7 | ESS11FullRaw$cgtsmok == 8 | ESS11FullRaw$cgtsmok == 9] <- NA
ESS11FullRaw$etfruit[ESS11FullRaw$etfruit == 77 | ESS11FullRaw$etfruit == 88 | ESS11FullRaw$etfruit == 99] <- NA
ESS11FullRaw$eatveg[ESS11FullRaw$eatveg == 77 | ESS11FullRaw$eatveg == 88 | ESS11FullRaw$eatveg == 99] <- NA
ESS11FullRaw$mbtru[ESS11FullRaw$mbtru == 7 | ESS11FullRaw$mbtru == 8 | ESS11FullRaw$mbtru == 9] <- NA
ESS11FullRaw$hincfel[ESS11FullRaw$hincfel == 7 | ESS11FullRaw$hincfel == 8 | ESS11FullRaw$hincfel == 9] <- NA
ESS11FullRaw$alcfreq[ESS11FullRaw$alcfreq == 77 | ESS11FullRaw$alcfreq == 88 | ESS11FullRaw$alcfreq == 99] <- NA
ESS11FullRaw$medtrun[ESS11FullRaw$medtrun == 7 | ESS11FullRaw$medtrun == 8 | ESS11FullRaw$medtrun == 9] <- NA
ESS11FullRaw$volunfp[ESS11FullRaw$volunfp == 7 | ESS11FullRaw$volunfp == 8 | ESS11FullRaw$volunfp == 9] <- NA
# Sex ----

ESS11Rec <- ESS11FullRaw %>% 
  mutate(
    gender = case_when( 
      gndr == 1 ~ 0,  # Male becomes 0
      gndr == 2 ~ 1   # Female becomes 1
    )
  )%>% 
  mutate(
    gender = set_value_labels(
      gender,
      labels = c("Male" = 0, "Female" = 1)
    )
  )

# Age ----

ESS11Rec <- ESS11Rec %>%
  mutate(
    age = 2024 - yrbrn
  ) 


# RegionWelfare ----

ESS11Rec <- ESS11Rec %>%
  mutate(regionWelfare = case_when(
    cntry %in% c("FI", "IS", "NO", "SE") ~ 1,  # Scandinavian
    cntry %in% c("GB", "IE") ~ 2,  # Anglo-Saxon
    cntry %in% c("AT", "BE", "CH", "DE", "FR","IL", "NL") ~ 3,  # Bismarkian
    cntry %in% c("BG", "EE", "HU", "HR", "LT", "LV","ME","PL", "RS", "SI", "SK", "UA") ~ 4,  #  Eastern European 
    cntry %in% c("ES", "GR", "IT", "PT","CY") ~ 5,  # Southern
    TRUE ~ 99  
  )) %>%
  mutate(regionWelfare = set_value_labels(
    regionWelfare,
    labels = c("Scandinavian" = 1, "Anglo-Saxon" = 2, "Mismarkian" = 3, "Eastern European " = 4,"Southern" = 5, "Other" = 99)
  ))

# Education ----

## eiscedV 
ESS11Rec <- ESS11Rec %>%
  mutate(
    eiscedV = case_when(
      eisced == 0 ~ eisced,
      eisced == 1 ~ eisced,
      eisced == 2 ~ eisced,
      eisced == 3 ~ eisced,
      eisced == 4 ~ 3,
      eisced == 5 ~ 4,
      eisced == 6 ~ 5,
      eisced == 7 ~ 5,
      eisced == 55 ~ NA_real_
    )
  ) %>% 
  mutate(
    eiscedV = set_value_labels(
      eiscedV,
      labels = c("Not possible to harmonise into ES-ISCED" = 0, "ES-ISCED I , less than lower secondary" = 1, 
                 "ES-ISCED II, lower secondary" = 2, "ES-ISCED III" = 3, "ISCED IV and ISCED v" = 4, "ISCED VI and over" = 5, 
                 "Other" = 55,"refusal" = 77, "Don't know" = 88, "No answer" = 99 ) 
    )
  )

# eisced2 
ESS11Rec <- ESS11Rec %>% 
  mutate( eisced2 = if_else(eiscedV < 3, 1, 0)) #eisced2 1 =	ES-ISCED IIIb, lower tier upper secondary OR lower, eisced2 0 = upper secondary or higher

# Health behaviours ----

## Smoking ----
ESS11Rec <- ESS11Rec %>% 
  mutate(
    cgtsmoker = case_when(
      cgtsmok == 1 ~ 5,
      cgtsmok == 2 ~ 4,
      cgtsmok == 3 ~ 3,
      cgtsmok == 4 ~ 2,
      cgtsmok == 5 ~ 1,
      cgtsmok == 6 ~ 0,
      cgtsmok == 7 ~ 7,
      cgtsmok == 8 ~ 8
    ) #cgtsmoke = 5 means i smoke daily
  ) %>% 
  mutate( 
    cgtsmoker = set_value_labels(
      cgtsmoker, 
      labels = c("never smoked" = 0, "few times" = 1, "not anymore" = 2, "not every day" = 3, "<= 9/day, daily" = 4, ">=10/day, daily" = 5, "Refusal" = 7, "Don't Know" = 8)
    )
  )

## Alcohol ----

ESS11Rec <- ESS11Rec %>% 
  mutate(
    alcoholfreq = case_when(
      alcfreq == 1 ~ 6,
      alcfreq == 2 ~ 5,
      alcfreq == 3 ~ 4,
      alcfreq == 4 ~ 3,
      alcfreq == 5 ~ 2,
      alcfreq == 6 ~ 1,
      alcfreq == 7 ~ 0,
      alcfreq == 77 ~ 77,
      alcfreq == 88 ~ 88) # alcoholfreq 6 means every day
  ) %>% 
  mutate(
    alcoholfreq = set_value_labels(
      alcoholfreq, labels = c("Never" = 0, "Less than once a month" = 1, "Once a month" = 2,"2-3 times a month" = 3, "Once a week" = 4, "Several times a week" = 5, "Every day" = 6, "Refusal" = 7, "Don't Know" = 8, "No answer" = 9)
    )
  )


## Physical activity ----
ESS11Rec <- ESS11Rec %>% 
  mutate(
    dosportr = case_when(
      dosprt == 0 ~ 7,
      dosprt == 1 ~ 6, 
      dosprt == 2 ~ 5,
      dosprt == 3 ~ 4,
      dosprt == 4 ~ 3,  
      dosprt == 5 ~ 2,  
      dosprt == 6 ~ 1 , 
      dosprt == 7 ~ 0 
    )
  ) # level of risk! 0 = made exercise everyday last week. 7 = never exercised


## Fruit and Vegetables ----

ESS11Rec <- ESS11Rec %>% 
  mutate(etfruitveg = (etfruit + eatveg)-2) # the higher value, the less fruit and veg being eaten



# Psychosocial ----  

## Control life ----
ESS11Rec <- ESS11Rec %>% 
  mutate(cntrlife = case_when(
    ctrlife == 0 ~ 10,
    ctrlife == 1 ~ 9,
    ctrlife == 2 ~ 8,
    ctrlife == 3 ~ 7,
    ctrlife == 4 ~ 6,
    ctrlife == 5 ~ 5,
    ctrlife == 6 ~ 4,
    ctrlife == 7 ~ 3,
    ctrlife == 8 ~ 2,
    ctrlife == 9 ~ 1,
    ctrlife == 10 ~ 0,
  )) # 0 is complete cntrl, 10 is no ctrl


## Life satisfaction ----
ESS11Rec <- ESS11Rec %>% 
  mutate(stflifer = case_when(
    stflife == 0 ~ 10,
    stflife == 1 ~ 9,
    stflife == 2 ~ 8,
    stflife == 3 ~ 7,
    stflife == 4 ~ 6,
    stflife == 5 ~ 5,
    stflife == 6 ~ 4,
    stflife == 7 ~ 3,
    stflife == 8 ~ 2,
    stflife == 9 ~ 1,
    stflife == 10 ~ 0,
  )) %>%  # 0 is Extremely satisfied, 10 is Extremely dissatisfied
  mutate(stflifer01 = stflifer/10)


## Childhood conflicts ----
ESS11Rec <- ESS11Rec %>% 
  mutate( #conflict
    conflpplh = case_when(
      cnfpplh == 1 ~ 4,
      cnfpplh == 2 ~ 3, 
      cnfpplh == 3 ~ 2,
      cnfpplh == 4 ~ 1,
      cnfpplh == 5 ~ 0,
    ) # conflpplh = 4 means always
  ) %>%   
  mutate(
    conflpplh = set_value_labels(
      conflpplh, labels = c("Never" = 0, "Hardly ever" = 1, "Sometimes" = 2,"Often" = 3,"Always" = 4, 
                            "Refusal" = 7, "Don't Know" = 8, "No answer" = 9)
    )
  ) 

# Employment conditions ----

## Unemployment ----
ESS11Rec <- ESS11Rec %>% 
  mutate(unemplD = if_else((uempli == 1)|(uempla == 1) , 1, 0)) # unemplD 1 means unemployed, either looking or not looking for a job


## Ergonomic ----
ESS11Rec <- ESS11Rec %>% 
  mutate(ergonomic = rowSums(across(
    c(jbexpmc, jbexpml, jbexpti, jbexpvi), 
    ~ .x == 1), 
    na.rm = TRUE)) 

## Material ----
ESS11Rec <- ESS11Rec %>% 
  mutate(material = rowSums(across(c(jbexebs, jbexecp, jbexera, jbexevc, jbexevh, jbexevl),
    ~ .x == 1), 
    na.rm = TRUE))

## Job control ---- 
ESS11Rec <- ESS11Rec %>% 
  mutate(wkorg = case_when(
    wkdcorga == 0 ~ 10,
    wkdcorga == 1 ~ 9,
    wkdcorga == 2 ~ 8,
    wkdcorga == 3 ~ 7,
    wkdcorga == 4 ~ 6,
    wkdcorga == 5 ~ 5,
    wkdcorga == 6 ~ 4,
    wkdcorga == 7 ~ 3,
    wkdcorga == 8 ~ 2,
    wkdcorga == 9 ~ 1,
    wkdcorga == 10 ~ 0,
  ))


## Job organization decisions ---- 
ESS11Rec <- ESS11Rec %>% 
  mutate(iorgactr = case_when(
    iorgact == 0 ~ 10,
    iorgact == 1 ~ 9,
    iorgact == 2 ~ 8,
    iorgact == 3 ~ 7,
    iorgact == 4 ~ 6,
    iorgact == 5 ~ 5,
    iorgact == 6 ~ 4,
    iorgact == 7 ~ 3,
    iorgact == 8 ~ 2,
    iorgact == 9 ~ 1,
    iorgact == 10 ~ 0,
    
  ))

# member of a trade union or similar ----
ESS11Rec <- ESS11Rec %>% 
  mutate (mbtruD = if_else(mbtru == 3 | mbtru == 2, 1, 0)) # no, or not currently


# Material conditions ----

## Financial difficulties growing up ----

ESS11Rec <- ESS11Rec %>% 
  mutate( 
    fndf = case_when(
      fnsdfml == 1 ~ 4,
      fnsdfml == 2 ~ 3, 
      fnsdfml == 3 ~ 2,
      fnsdfml == 4 ~ 1,
      fnsdfml == 5 ~ 0,
    ) # fndf 4 means 	Always	inancial difficulties growing up
  ) %>% 
  mutate(
    fndf = set_value_labels(
      fndf, labels = c("Never" = 0, "Hardly ever" = 1, "Sometimes" = 2,"Often" = 3,"Always" = 4, "Refusal" = 7, "Don't Know" = 8, "No answer" = 9)
    )
  )


## Housing problems ----
ESS11Rec <- ESS11Rec %>% 
  mutate(housing = rowSums(across(
    c(paccmoro, paccdwlr, pacclift, paccnbsh, paccocrw, paccxhoc, paccnois, paccinro), 
    ~ .x == 1), 
    na.rm = TRUE)) 


# Social Capital ----

## Meeting people ----
ESS11Rec <- ESS11Rec %>% 
  mutate(socialmeet = case_when(
    sclmeet == 1 ~ 6,
    sclmeet == 2 ~ 5,
    sclmeet == 3 ~ 4,
    sclmeet == 4 ~ 3,
    sclmeet == 5 ~ 2,
    sclmeet == 6 ~ 1,
    sclmeet == 7 ~ 0,
  )) # risk! socialmeet 0 = means everyday meetings, sclmeet 6 == never

## people with whom to talk ----
ESS11Rec <- ESS11Rec %>% 
  mutate(inperdisc = case_when(
    inprdsc == 0 ~ 6,
    inprdsc == 1 ~ 5,
    inprdsc == 2 ~ 4,
    inprdsc == 3 ~ 3,
    inprdsc == 4 ~ 2,
    inprdsc == 5 ~ 1,
    inprdsc == 6 ~ 0,
  )) # risk!  inperdisc 0 >= 10 ppl to talk, inperdisc 6 = none


## volunteered  ----
ESS11Rec <- ESS11Rec %>% 
  mutate (volunfpr = if_else(volunfp == 2, 1, 0))


## People can be trusted ----
ESS11Rec <- ESS11Rec %>% 
  mutate(ppltrust = case_when(
    ppltrst == 0 ~ 10,
    ppltrst == 1 ~ 9,
    ppltrst == 2 ~ 8,
    ppltrst == 3 ~ 7,
    ppltrst == 4 ~ 6,
    ppltrst == 5 ~ 5,
    ppltrst == 6 ~ 4,
    ppltrst == 7 ~ 3,
    ppltrst == 8 ~ 2,
    ppltrst == 9 ~ 1,
    ppltrst == 10 ~ 0,
  ))  # risk! ppltrust 0 is 	Most people can be trusted, 10 is You can't be too careful


## Politicians can be trusted ----
ESS11Rec <- ESS11Rec %>% 
  mutate(trustplt = case_when(
    trstplt == 0 ~ 10,
    trstplt == 1 ~ 9,
    trstplt == 2 ~ 8,
    trstplt == 3 ~ 7,
    trstplt == 4 ~ 6,
    trstplt == 5 ~ 5,
    trstplt == 6 ~ 4,
    trstplt == 7 ~ 3,
    trstplt == 8 ~ 2,
    trstplt == 9 ~ 1,
    trstplt == 10 ~ 0,
  )) 


# Healthcare ----

## Unpaid care ---- 
ESS11Rec <- ESS11Rec %>% 
  mutate(hlpfmhrr = case_when( ## helps, and does it more than 10hrs/week
    hlpfmhr == 1 ~ hlpfmhr,
    hlpfmhr == 2 ~ hlpfmhr,
    hlpfmhr == 3 ~ hlpfmhr,
    hlpfmhr == 4 ~ hlpfmhr,
    hlpfmhr == 5 ~ hlpfmhr,
    hlpfmhr == 6 ~ hlpfmhr,
    hlpfmhr == 55 ~ 0,
      ))

ESS11Rec <- ESS11Rec %>% 
  mutate(
    hlpfmhrrcomb = case_when(
      hlpfmly == 2 ~ 0,
      hlpfmhrr == 0 ~ hlpfmhrr, 
      hlpfmhrr > 0 ~ hlpfmhrr )
  ) 

## HC unmet need ----
ESS11Rec <- ESS11Rec %>% 
  mutate(
    medtrun2 = case_when(
      medtrun == 1 ~ 1,
      medtrun == 2 ~ 0, 
      medtrun == 7 ~ 7,
      medtrun == 8 ~ 8,
    ) 
  ) %>% 
  mutate(
    medtrun2 = set_value_labels(
      medtrun2, labels = c("unmet need" = 1, "No unmet need" = 0,  "Refusal" = 7, "Don't Know" = 8)
    )
  )



## General Practitioner ----
ESS11Rec <- ESS11Rec %>% 
  mutate(dshltgpr = if_else(dshltgp == 0, 1, 0)) #not discussed, dshltgpr = 1, discussed = 0

## Medical specialist ----
ESS11Rec <- ESS11Rec %>% 
  mutate(dshltmsr = if_else(dshltms == 0, 1, 0))



# Identification number per row ----
ESS11Rec <- ESS11Rec %>% 
  mutate(id_row = row_number())

```

```{r}
# Deletion of Israel
ESS11RecEu <- ESS11Rec %>% 
  filter(cntry != "IL")

# Produce a vector of all countries
country_list <- unique(ESS11RecEu$cntry)

# Prepare region and regionWelfare datasets to join later with the Block_scores datasets 
country_regionWelfare <- ESS11RecEu %>%
  dplyr::select(cntry, regionWelfare) %>%
  distinct()

WR_list <- country_regionWelfare %>% 
  distinct(regionWelfare) %>% 
  filter(regionWelfare != 99) %>%
  pull(regionWelfare) %>%
  sort()
WR_labels <- c("1" = "Scandinavian", "2" = "Anglo-Saxon", "3" = "Bismarckian", 
               "4" = "Eastern European", "5" = "Southern")
```


# Statistical analysis

## Additional libraries for statistical analysis

```{r}
library(survey)
library(haven)
library(dplyr)
library(ggplot2)
library(tidyr)
library(lme4)
library(readxl)
library(labelled)
library(openxlsx)
library(ltm)
library(psych)
library(ggplot2)
library(forcats)
library(fmsb)
library(janitor)
library(effectsize)
library(car)
library(dendextend)
```

## PCA: Composite risk score

### Preparations
```{r}
# Create a DataSet 
pca_variables <- c("id_row", "cntry", "regionWelfare", "psu", "stratum", "gender", "pspwght", "anweight", "eisced2", "material", "ergonomic", "wkorg", "iorgactr", "unemplD", "mbtruD", "housing", "fndf", "hincfel", "medtrun2", "hlpfmhrrcomb", "dshltgpr","dshltmsr" , "etfruitveg", "dosportr", "cgtsmoker", "alcoholfreq", "cntrlife", "stflifer","conflpplh", "socialmeet", "inperdisc", "ppltrust", "trustplt", "volunfpr")

# Check NAs
colSums(is.na(ESS11RecEu[,pca_variables])) 

# Make dataset
pca_df <- ESS11RecEu %>% 
  dplyr::select(all_of(pca_variables))

# Include sex as factor
pca_df <- pca_df %>% 
  mutate(sex = factor(gender, levels = c(0,1), labels = c("Men", "Women")))
```


  
### Education
```{r}
#Step 0: Make a df only for the subset of variables (to drop NAs)
index_df_edu <- pca_df %>% 
  drop_na(c("id_row","eisced2") ) 


#Step 1: Calculate Z-Scores = standardized comparative indicator
index_df_edu_scores <- index_df_edu %>%
  mutate(
    index_edu = as.numeric(scale(eisced2))
  )

#Step 2: Define survey design
index_design_edu <- svydesign(
  ids = ~psu,
  strata = ~stratum,
  weights = ~pspwght,
  data = index_df_edu_scores,
  nest = TRUE
)
options(survey.lonely.psu = "adjust")

# Step 3: Estimate means with confidence intervals

#By sex
edu_index_bysex <- svyby( 
  ~index_edu,
  ~cntry + sex, 
  index_design_edu,
  svymean,
  na.rm = TRUE,
  vartype = c("se", "ci")
)

#pooled sex
edu_index_pooledsex <- svyby(
  ~index_edu,
  ~cntry,
  index_design_edu,
  svymean,
  na.rm = TRUE,
  vartype = c("se", "ci")
) %>% 
  as.data.frame() %>% 
  mutate(sex = "All")  


# Europe pooled (mean of countries)
edu_index_pooledCountry <- mean(edu_index_pooledsex$index_edu, na.rm = TRUE)

# WR pooled (mean of countries)
edu_index_pooledWR <-  edu_index_pooledsex %>%
  left_join(country_regionWelfare, by = "cntry") %>% 
  group_by(regionWelfare) %>%
  summarise(
    wr_mean_edu = mean(index_edu, na.rm = TRUE),
    .groups = "drop"
  )

# Step 4: Bind analyses, and add region info back in 
edu_index_score <- bind_rows(
  edu_index_pooledsex %>%  dplyr::select(cntry, index_edu, se, ci_l, ci_u, sex),
  edu_index_bysex %>% dplyr::select(cntry, index_edu, se, ci_l, ci_u, sex)
)

edu_index_score <- edu_index_score %>% 
  left_join(country_regionWelfare, by = "cntry")
```

### Employment Conditions

```{r}
#Step 0: Make a df only for the subset of variables (to drop NAs)
pca_df_work <- pca_df %>% 
  drop_na(c( "unemplD", "mbtruD", "material","ergonomic","wkorg","iorgactr") )

# Calculate a manual reference Risk score to know the direction of risk (and correct the PCA sign if needed)
risk_ref_work <- pca_df_work %>% 
  mutate(across(
    c("unemplD", "mbtruD", "material","ergonomic","wkorg","iorgactr"),
    ~ (.-min(., na.rm=TRUE)) / (max(., na.rm=TRUE) - min(., na.rm=TRUE))
  )) %>% 
  transmute(risk_ref_psych = rowMeans(across(c("unemplD", "mbtruD", "material","ergonomic","wkorg","iorgactr")), na.rm = TRUE)) %>% 
  pull()


# Step 1: Run PCA
pca_work <- prcomp( pca_df_work[, c("material","ergonomic","unemplD","wkorg","iorgactr","mbtruD")], 
                    scale. = TRUE
)


# Check if sign is aligned with risk reference 
cor_work <- cor(pca_work$x[,1], risk_ref_work, use = "complete.obs")

if(cor_work < 0){
  pca_work$x[,1] <- -pca_work$x[,1]
  pca_work$rotation[,1] <- -pca_work$rotation[,1]
}

print(pca_work)

# Merge the result of the PCA to the initial pca_df 
pca_df_work_scores <- pca_df_work %>%
  mutate(pc1_work = pca_work$x[,1])


# Step 2: Define survey design to be able to do survey-weighed calculations later on
pca_design_work <- svydesign(
  ids = ~psu,
  strata = ~stratum,
  weights = ~pspwght,
  data = pca_df_work_scores,  
  nest = TRUE
)
options(survey.lonely.psu = "adjust")


# Step 3: Estimate means with confidence intervals

#By sex
work_pca_bysex <- svyby( 
  ~pc1_work, 
  ~cntry + sex, 
  pca_design_work,
  svymean,
  na.rm = TRUE,
  vartype = c("se", "ci")
)

#pooled sex 
work_pca_pooledsex <- svyby(
  ~pc1_work,
  ~cntry,
  pca_design_work,
  svymean,
  na.rm = TRUE,
  vartype = c("se", "ci")
) %>% 
  as.data.frame() %>% 
  mutate(sex = "All") 


# Europe pooled (mean of countries)
work_pca_pooledCountry <- mean(work_pca_pooledsex$pc1_work, na.rm = TRUE)

# WR pooled (mean of countries) 
work_pca_pooledWR <-  work_pca_pooledsex %>%
  left_join(country_regionWelfare, by = "cntry") %>% 
  group_by(regionWelfare) %>%
  summarise(
    wr_mean_work = mean(pc1_work, na.rm = TRUE),
    .groups = "drop"
  )

# Step 4: Bind analyses, and add region info back in (once per country)
work_pca_score <- bind_rows(
  work_pca_pooledsex %>%  dplyr::select(cntry, pc1_work, se, ci_l, ci_u, sex),
  work_pca_bysex %>% dplyr::select(cntry, pc1_work, se, ci_l, ci_u, sex)
)


work_pca_score <- work_pca_score %>% 
  left_join(country_regionWelfare, by = "cntry")

```

### Material circumstances

```{r}
#Step 0: Make a df only for the subset of variables (to drop NAs)
pca_df_mat <- pca_df %>% 
  drop_na(c("id_row","housing", "fndf", "hincfel") )


# Calculate a manual reference Risk score to know the direction of risk (to correct the PCA sign if needed)
risk_ref_mat <- pca_df_mat %>% 
  mutate(across(
    c("housing", "fndf", "hincfel"),
    ~ (.-min(., na.rm=TRUE)) / (max(., na.rm=TRUE) - min(., na.rm=TRUE))
  )) %>% 
  transmute(risk_ref_psych = rowMeans(across(c("housing", "fndf", "hincfel")), na.rm = TRUE)) %>% 
  pull()


# Step 1: Run PCA 

pca_mat <- prcomp(pca_df_mat[, c("housing", "fndf", "hincfel")], 
                  scale. = TRUE
)

print(pca_mat)

# Check if sign is aligned with risk reference

cor_mat <- cor(pca_mat$x[,1], risk_ref_mat, use = "complete.obs")

if(cor_mat < 0){
  pca_mat$x[,1] <- -pca_mat$x[,1]
  pca_mat$rotation[,1] <- -pca_mat$rotation[,1]
}
print(pca_mat)

# Merge the result of the PCA to the initial pca_df
pca_df_mat_scores <- pca_df_mat %>% 
  mutate(pc1_mat = pca_mat$x[,1]) 


# Step 2: Define survey design 
pca_design_mat <- svydesign(
  ids = ~psu,
  strata = ~stratum,
  weights = ~pspwght,
  data = pca_df_mat_scores,
  nest = TRUE
)
options(survey.lonely.psu = "adjust")


# Step 3: Estimate means with confidence intervals

#By sex
mat_pca_bysex <- svyby( 
  ~pc1_mat,
  ~cntry + sex, 
  pca_design_mat,
  svymean,
  na.rm = TRUE,
  vartype = c("se", "ci")
)

#pooled sex 
mat_pca_pooledsex <- svyby(
  ~pc1_mat,
  ~cntry,
  pca_design_mat,
  svymean,
  na.rm = TRUE,
  vartype = c("se", "ci")
) %>% 
  as.data.frame() %>% 
  mutate(sex = "All")   


# Europe pooled (mean of countries)

mat_pca_pooledCountry <- mean(mat_pca_pooledsex$pc1_mat, na.rm = TRUE)

# WR pooled (mean of countries)

mat_pca_pooledWR <-  mat_pca_pooledsex %>%
  left_join(country_regionWelfare, by = "cntry") %>% 
  group_by(regionWelfare) %>%
  summarise(
    wr_mean_mat = mean(pc1_mat, na.rm = TRUE),
    .groups = "drop"
  )

# Step 4: Bind analyses, and add region info back in (once per country) 

mat_pca_score <- bind_rows(
  mat_pca_pooledsex %>%  dplyr::select(cntry, pc1_mat, se, ci_l, ci_u, sex),
  mat_pca_bysex %>% dplyr::select(cntry, pc1_mat, se, ci_l, ci_u, sex)
)

mat_pca_score <- mat_pca_score %>% 
  left_join(country_regionWelfare, by = "cntry")

```

### HC access

```{r}
# Step 0: Make a df only for the subset of variables (to drop NAs)
pca_df_hc <- pca_df %>% 
  drop_na(c("medtrun2", "hlpfmhrrcomb", "dshltmsr", "dshltgpr") )

# Calculate a manual reference Risk score to know the direction of risk (and correct the PCA sign if needed)
risk_ref_hc <- pca_df_hc %>% 
  mutate(across(c("medtrun2", "dshltmsr"),
                ~ (.-min(., na.rm=TRUE)) / (max(., na.rm=TRUE) - min(., na.rm=TRUE))
  )) %>% 
  transmute(risk_ref_hc = rowMeans(across(c("medtrun2")), na.rm = TRUE)) %>% 
  pull()


# Step 1: Run PCA 
pca_hc <- prcomp(pca_df_hc[, c( "medtrun2", "hlpfmhrrcomb", "dshltmsr", "dshltgpr")],
                 scale. = TRUE) 

print(pca_hc)

#Check if sign is aligned with risk reference
cor_hc <- cor(pca_hc$x[,1], risk_ref_hc, use = "complete.obs")

if(cor_hc < 0){
  pca_hc$x[,1] <- -pca_hc$x[,1]
  pca_hc$rotation[,1] <- -pca_hc$rotation[,1]
}
print(pca_hc)

# Merge the result of the PCA to the initial pca_df
pca_df_hc_scores <- pca_df_hc %>% 
  mutate(pc1_hc = pca_hc$x[,1]) 

# Step 2: Define survey design
pca_design_hc <- svydesign(
  ids = ~psu,
  strata = ~stratum,
  weights = ~pspwght,
  data = pca_df_hc_scores,
  nest = TRUE
)

# Step 3: Estimate means with confidence intervals

#By sex
hc_pca_bysex <- svyby( 
  ~pc1_hc,
  ~cntry + sex, 
  pca_design_hc,
  svymean,
  na.rm = TRUE,
  vartype = c("se", "ci")
)

#pooled sex 
hc_pca_pooledsex <- svyby(
  ~pc1_hc,
  ~cntry,
  pca_design_hc,
  svymean,
  na.rm = TRUE,
  vartype = c("se", "ci")
) %>% 
  as.data.frame() %>% 
  mutate(sex = "All")   


# Europe pooled (mean of countries)

hc_pca_pooledCountry <- mean(hc_pca_pooledsex$pc1_hc, na.rm = TRUE)

# WR pooled (mean of countries)

hc_pca_pooledWR <-  hc_pca_pooledsex %>%
  left_join(country_regionWelfare, by = "cntry") %>% 
  group_by(regionWelfare) %>%
  summarise(
    wr_mean_hc = mean(pc1_hc, na.rm = TRUE),
    .groups = "drop"
  )

# Step 4: Bind analyses, and add region info back in (once per country) 
hc_pca_score <- bind_rows(
  hc_pca_pooledsex %>%  dplyr::select(cntry, pc1_hc, se, ci_l, ci_u, sex),
  hc_pca_bysex %>% dplyr::select(cntry, pc1_hc, se, ci_l, ci_u, sex)
)

hc_pca_score <- hc_pca_score %>% 
  left_join(country_regionWelfare, by = "cntry")
```

### Behavioural

```{r}
# Step 0: Make a df only for the subset of variables (to drop NAs)
pca_df_behav <- pca_df %>% 
  drop_na(c("id_row","etfruitveg", "dosportr", "cgtsmoker", "alcoholfreq"
  ) )

# Calculate a manual reference Risk score to know the direction of risk (and correct the PCA sign if needed);

risk_ref_behav <- pca_df_behav %>% 
  mutate(across(
    c("etfruitveg", "dosportr", "cgtsmoker", "alcoholfreq"),
    ~ (.-min(., na.rm=TRUE)) / (max(., na.rm=TRUE) - min(., na.rm=TRUE))
  )) %>% 
  transmute(risk_ref_behav = rowMeans(across(c("etfruitveg", "dosportr", "cgtsmoker", "alcoholfreq")), na.rm = TRUE)) %>% ## changed!!! , "alcoholbngcomb"
  pull()


# Step 1: Run PCA 

pca_behav <- prcomp(pca_df_behav[, c("etfruitveg", "dosportr", "cgtsmoker", "alcoholfreq")], 
                    scale. = TRUE) 

print(pca_behav)

#Check if sign is aligned with risk reference
cor_behav <- cor(pca_behav$x[,1], risk_ref_behav, use = "complete.obs")

if(cor_behav < 0){
  pca_behav$x[,1] <- -pca_behav$x[,1]
  pca_behav$rotation[,1] <- -pca_behav$rotation[,1]
}
print(pca_behav)


# Merge the result of the PCA to the initial pca_df

pca_df_behav_scores <- pca_df_behav %>% 
  mutate(pc1_behav = pca_behav$x[,1]) 


# Step 2: Define survey design
pca_design_behav <- svydesign(
  ids = ~psu,
  strata = ~stratum,
  weights = ~pspwght,
  data = pca_df_behav_scores,
  nest = TRUE
)


# Step 3: Estimate means with confidence intervals

#By sex
behav_pca_bysex <- svyby( 
  ~pc1_behav,
  ~cntry + sex, 
  pca_design_behav,
  svymean,
  na.rm = TRUE,
  vartype = c("se", "ci")
)

#pooled sex 
behav_pca_pooledsex <- svyby(
  ~pc1_behav,
  ~cntry,
  pca_design_behav,
  svymean,
  na.rm = TRUE,
  vartype = c("se", "ci")
) %>% 
  as.data.frame() %>% 
  mutate(sex = "All")   


# Europe pooled (mean of countries)

behav_pca_pooledCountry <- mean(behav_pca_pooledsex$pc1_behav, na.rm = TRUE)

# WR pooled (mean of countries)

behav_pca_pooledWR <-  behav_pca_pooledsex %>%
  left_join(country_regionWelfare, by = "cntry") %>% 
  group_by(regionWelfare) %>%
  summarise(
    wr_mean_behav = mean(pc1_behav, na.rm = TRUE),
    .groups = "drop"
  )


# Step 4: Bind analyses 

behav_pca_score <- bind_rows(
  behav_pca_pooledsex %>%  dplyr::select(cntry, pc1_behav, se, ci_l, ci_u, sex),
  behav_pca_bysex %>% dplyr::select(cntry, pc1_behav, se, ci_l, ci_u, sex)
)


behav_pca_score <- behav_pca_score %>% 
  left_join(country_regionWelfare, by = "cntry")

```

### Psychosocial

```{r}

# Step 0: Make a df only for the subset of variables (to drop NAs)
pca_df_psych <- pca_df %>% 
  drop_na(c("id_row","conflpplh", "cntrlife", "stflifer"))


# Calculate a manual reference Risk score to know the direction of risk (and correct the PCA sign if needed)

risk_ref_psych <- pca_df_psych %>% 
  mutate(across(
    c("conflpplh", "cntrlife", "stflifer"),
    ~ (.-min(., na.rm=TRUE)) / (max(., na.rm=TRUE) - min(., na.rm=TRUE))
  )) %>% 
  transmute(risk_ref_psych = rowMeans(across(c("conflpplh", "cntrlife", "stflifer")), na.rm = TRUE)) %>% 
  pull()


# Step 1: Run PCA 
pca_psych <- prcomp(pca_df_psych [, c( "conflpplh","cntrlife","stflifer")], 
                    scale. = TRUE)

print(pca_psych)

#Check if sign is aligned with risk reference
cor_psych <- cor(pca_psych$x[,1], risk_ref_psych, use = "complete.obs")

if(cor_psych < 0){
  pca_psych$x[,1] <- -pca_psych$x[,1]
  pca_psych$rotation[,1] <- -pca_psych$rotation[,1]
}
print(pca_psych)

# Merge the result of the PCA to the initial pca_df

pca_df_psych_scores <- pca_df_psych %>% 
  mutate(pc1_psych = pca_psych$x[,1]) 


# Step 2: Define survey design
pca_design_psych <- svydesign(
  ids = ~psu,
  strata = ~stratum,
  weights = ~pspwght,
  data = pca_df_psych_scores,
  nest = TRUE
)

# Step 3: Estimate means with confidence intervals

#By sex
psych_pca_bysex <- svyby( 
  ~pc1_psych,
  ~cntry + sex, 
  pca_design_psych,
  svymean,
  na.rm = TRUE,
  vartype = c("se", "ci")
)

#pooled sex 
psych_pca_pooledsex <- svyby(
  ~pc1_psych,
  ~cntry,
  pca_design_psych,
  svymean,
  na.rm = TRUE,
  vartype = c("se", "ci")
) %>% 
  as.data.frame() %>% 
  mutate(sex = "All")   # add label so we can row-bind

# Europe pooled (mean of countries)

psych_pca_pooledCountry <- mean(psych_pca_pooledsex$pc1_psych, na.rm = TRUE)

# WR pooled (mean of countries)

psych_pca_pooledWR <-  psych_pca_pooledsex %>%
  left_join(country_regionWelfare, by = "cntry") %>% 
  group_by(regionWelfare) %>%
  summarise(
    wr_mean_psych = mean(pc1_psych, na.rm = TRUE),
    .groups = "drop"
  )

# Step 4: Bind analyses, and add region info back in (once per country) 

psych_pca_score <- bind_rows(
  psych_pca_pooledsex %>%  dplyr::select(cntry, pc1_psych, se, ci_l, ci_u, sex),
  psych_pca_bysex %>% dplyr::select(cntry, pc1_psych, se, ci_l, ci_u, sex)
)


psych_pca_score <- psych_pca_score %>% 
  left_join(country_regionWelfare, by = "cntry")

```

### Social Capital

```{r}
# Step 0: Make a df only for the subset of variables (to drop NAs)
pca_df_sc <- pca_df %>% 
  drop_na(c("id_row","socialmeet", "inperdisc", "ppltrust", "trustplt", "volunfpr") )


# Calculate a manual reference Risk score to know the direction of risk (and correct the PCA sign if needed)

risk_ref_sc <- pca_df_sc %>% 
  mutate(across(
    c("socialmeet", "inperdisc", "ppltrust", "trustplt", "volunfpr"),
    ~ (.-min(., na.rm=TRUE)) / (max(., na.rm=TRUE) - min(., na.rm=TRUE))
  )) %>% 
  transmute(risk_ref_sc = rowMeans(across(c("socialmeet", "inperdisc", "ppltrust", "trustplt", "volunfpr")), na.rm = TRUE)) %>% 
  pull()



# Step 1: Run PCA 

pca_sc <- prcomp(pca_df_sc[, c( "socialmeet", "inperdisc", "ppltrust", "trustplt","volunfpr")], 
                 scale. = TRUE) # Run PCA on selected variables

print(pca_sc)

# Check if sign is aligned with risk reference
cor_sc <- cor(pca_sc$x[,1], risk_ref_sc, use = "complete.obs")

if(cor_sc < 0){
  pca_sc$x[,1] <- -pca_sc$x[,1]
  pca_sc$rotation[,1] <- -pca_sc$rotation[,1]
}
print(pca_sc)


# Merge pca_df_hc_scores to have the individual PC
pca_df_sc_scores <- pca_df_sc %>% 
  mutate(pc1_sc = pca_sc$x[,1]) 


# Step 2: Define survey design
pca_design_sc <- svydesign(
  ids = ~psu,
  strata = ~stratum,
  weights = ~pspwght,
  data = pca_df_sc_scores,
  nest = TRUE
)


# Step 3: Estimate means with confidence intervals

#By sex
sc_pca_bysex <- svyby( 
  ~pc1_sc,
  ~cntry + sex, 
  pca_design_sc,
  svymean,
  na.rm = TRUE,
  vartype = c("se", "ci")
)

#pooled sex 
sc_pca_pooledsex <- svyby(
  ~pc1_sc,
  ~cntry,
  pca_design_sc,
  svymean,
  na.rm = TRUE,
  vartype = c("se", "ci")
) %>% 
  as.data.frame() %>% 
  mutate(sex = "All")   # add label so we can row-bind


# Europe pooled (mean of countries)

sc_pca_pooledCountry <- mean(sc_pca_pooledsex$pc1_sc, na.rm = TRUE)

# WR pooled (mean of countries)

sc_pca_pooledWR <-  sc_pca_pooledsex %>%
  left_join(country_regionWelfare, by = "cntry") %>% 
  group_by(regionWelfare) %>%
  summarise(
    wr_mean_sc = mean(pc1_sc, na.rm = TRUE),
    .groups = "drop"
  )

# Step 4: Bind analyses, and add region info back in (once per country) 

sc_pca_score <- bind_rows(
  sc_pca_pooledsex %>%  dplyr::select(cntry, pc1_sc, se, ci_l, ci_u, sex),
  sc_pca_bysex %>% dplyr::select(cntry, pc1_sc, se, ci_l, ci_u, sex)
)

sc_pca_score <- sc_pca_score %>% 
  left_join(country_regionWelfare, by = "cntry")
```

## PLOTS: PCA Composite Scores (Supplementary material)

### Education

```{r}
# Get positions of countries per regime
regime_spans <- edu_index_score %>% 
  arrange(regionWelfare, cntry) %>%
  distinct(cntry, regionWelfare) %>%
  mutate(cntry_id = as.numeric(factor(cntry, levels = unique(cntry)))) %>%
  group_by(regionWelfare) %>%
  summarise(x_start = min(cntry_id), x_end = max(cntry_id), .groups = "drop")

# Get positions of the lines for regime averages 
regime_spans_eduWR <- regime_spans %>% 
  left_join(edu_index_pooledWR, by = "regionWelfare")

#plot
ggplot(edu_index_score %>%  
         arrange(regionWelfare) %>% 
         mutate(cntry = factor(cntry, levels = unique(cntry))),
       aes(x = cntry, y = index_edu, ymin = ci_l, ymax = ci_u, color = sex)) + 
  geom_pointrange(size = 0.28,
                  aes(alpha = ifelse(sex == "All", 1, 0.4)) ) +
  scale_color_manual(
    values = c("All" = "black", "Men" = "#1F77B4", "Women" = "#E69F00"),
    breaks = c("All", "Men", "Women")
  ) +
  scale_alpha_identity() + 
  geom_hline(yintercept = edu_index_pooledCountry, linetype = "dotted", color = "black") +
  geom_segment(
    data = regime_spans_eduWR,
    aes(x = x_start, xend = x_end, y = wr_mean_edu, yend = wr_mean_edu),
    inherit.aes = F, linetype = "11", linewidth = 0.7, color = "blue")  +
  geom_segment( 
    data = regime_spans,
    aes(x = x_start, xend = x_end, y = -1.55, yend = -1.55), 
    inherit.aes = FALSE, linewidth = 1.2, color = "grey40"
  ) +
  
  coord_cartesian(ylim = c(-1.5, 1.5)) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Education disadvantage",
    y = "Education Score (standardized index)"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.x = element_blank(),   
        plot.background   = element_rect(fill = "white", color = NA),
        panel.background  = element_rect(fill = "white", color = NA),
        legend.position = "none")


ggsave("Figures\\CompositeScores\\eduWR_psp.png", width = 6, height = 4, dpi = 300)
```

### Employment Conditions

```{r}
# Get positions of countries per regime
regime_spans <- work_pca_score %>%
  arrange(regionWelfare, cntry) %>%
  distinct(cntry, regionWelfare) %>%
  mutate(cntry_id = as.numeric(factor(cntry, levels = unique(cntry)))) %>%
  group_by(regionWelfare) %>%
  summarise(x_start = min(cntry_id), x_end = max(cntry_id), .groups = "drop")


# Get positions of the lines for regime averages 
regime_spans_workWR <- regime_spans %>% 
  left_join(work_pca_pooledWR, by = "regionWelfare")


#plot

ggplot(work_pca_score %>% 
         arrange(regionWelfare) %>% 
         mutate(cntry = factor(cntry, levels = unique(cntry))),
       aes(x = cntry, y = pc1_work, ymin = ci_l, ymax = ci_u, color = sex)) + 
  geom_pointrange(size = 0.28,
                  aes(alpha = ifelse(sex == "All", 1, 0.4)) ) +
  scale_color_manual(
    values = c("All" = "black", "Men" = "#1F77B4", "Women" = "#E69F00"),
    breaks = c("All", "Men", "Women")
  ) +
  scale_alpha_identity() +  
  geom_hline(yintercept = work_pca_pooledCountry, linetype = "dotted", color = "black") +
  geom_segment(
    data = regime_spans_workWR,
    aes(x = x_start, xend = x_end, y = wr_mean_work, yend = wr_mean_work),
    inherit.aes = F, linetype = "11", linewidth = 0.7, color = "blue")  +
  geom_segment( 
    data = regime_spans,
    aes(x = x_start, xend = x_end, y = -1.55, yend = -1.55), 
    inherit.aes = FALSE,
    size = 1.2,
    color = "grey40"
  ) +
  coord_cartesian(ylim = c(-2, 2)) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Employment disadvantage",
    y = "Composite Employment Score (PCA)"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.x = element_blank(),
        plot.background   = element_rect(fill = "white", color = NA),
        panel.background  = element_rect(fill = "white", color = NA),
        legend.position = "none")



ggsave("Figures\\CompositeScores\\workWR_psp.png", width = 6, height = 4, dpi = 300)
```

### Material

```{r}
# Get positions of countries per regime
regime_spans <- mat_pca_score %>%
  arrange(regionWelfare, cntry) %>%
  distinct(cntry, regionWelfare) %>%
  mutate(cntry_id = as.numeric(factor(cntry, levels = unique(cntry)))) %>%
  group_by(regionWelfare) %>%
  summarise(x_start = min(cntry_id), x_end = max(cntry_id), .groups = "drop")

# Get positions of the lines for regime averages 
regime_spans_matWR <- regime_spans %>% 
  left_join(mat_pca_pooledWR, by = "regionWelfare")


#plot
ggplot(mat_pca_score %>% 
         arrange(regionWelfare) %>% 
         mutate(cntry = factor(cntry, levels = unique(cntry))),
       aes(x = cntry, y = pc1_mat, ymin = ci_l, ymax = ci_u, color = sex)) + 
  geom_pointrange(size = 0.28,
                  aes(alpha = ifelse(sex == "All", 1, 0.4)) ) +
  scale_color_manual(
    values = c("All" = "black", "Men" = "#1F77B4", "Women" = "#E69F00"),
    breaks = c("All", "Men", "Women")
  ) +
  scale_alpha_identity() +  
  geom_hline(yintercept = mat_pca_pooledCountry, linetype = "dotted", color = "black") +
  geom_segment(
    data = regime_spans_matWR,
    aes(x = x_start, xend = x_end, y = wr_mean_mat, yend = wr_mean_mat),
    inherit.aes = F, linetype = "11", linewidth = 0.7, color = "blue")  +
  geom_segment( 
    data = regime_spans,
    aes(x = x_start, xend = x_end, y = -1.55, yend = -1.55), 
    inherit.aes = FALSE,
    linewidth = 1.2,
    color = "grey40"
  ) +
  coord_cartesian(ylim = c(-1.5, 1.5)) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Comparison of Material risk",
    y = "Composite Material Score (PCA)"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.x = element_blank(),
        plot.background = element_rect(fill = "white", color = NA),
        panel.background  = element_rect(fill = "white", color = NA),
        legend.position = "none")

ggsave("Figures\\CompositeScores\\matWR_psp.png", width = 6, height = 4, dpi = 300)
```

### HC access

```{r}

# Get positions of countries per regime
regime_spans <- hc_pca_score %>%
  arrange(regionWelfare, cntry) %>%
  distinct(cntry, regionWelfare) %>%
  mutate(cntry_id = as.numeric(factor(cntry, levels = unique(cntry)))) %>%
  group_by(regionWelfare) %>%
  summarise(x_start = min(cntry_id), x_end = max(cntry_id), .groups = "drop")

# Get positions of the lines for regime averages 
regime_spans_hcWR <- regime_spans %>% 
  left_join(hc_pca_pooledWR, by = "regionWelfare")


#plot
ggplot(hc_pca_score %>% 
         arrange(regionWelfare) %>% 
         mutate(cntry = factor(cntry, levels = unique(cntry))),
       aes(x = cntry, y = pc1_hc, ymin = ci_l, ymax = ci_u, color = sex)) + 
  geom_pointrange(size = 0.28,
                  aes(alpha = ifelse(sex == "All", 1, 0.4)) ) +
  scale_color_manual(
    values = c("All" = "black", "Men" = "#1F77B4", "Women" = "#E69F00"),
    breaks = c("All", "Men", "Women")
  ) +
  scale_alpha_identity() +  
  geom_hline(yintercept = hc_pca_pooledCountry, linetype = "dotted", color = "black") +
  geom_segment(
    data = regime_spans_hcWR,
    aes(x = x_start, xend = x_end, y = wr_mean_hc, yend = wr_mean_hc),
    inherit.aes = F, linetype = "11", linewidth = 0.7, color = "blue")  +
  geom_segment( 
    data = regime_spans,
    aes(x = x_start, xend = x_end, y = -1.55, yend = -1.55), 
    inherit.aes = FALSE,
    size = 1.2,
    color = "grey40"
  ) +
  coord_cartesian(ylim = c(-1.5, 1.5)) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Healthcare access disadvantage", 
    y = "Composite Healthcare access Score (PCA)"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.x = element_blank(),
        plot.background   = element_rect(fill = "white", color = NA),
        panel.background  = element_rect(fill = "white", color = NA),
        legend.position = "none")


ggsave("Figures\\CompositeScores\\HCWR_psp.png", width = 6, height = 4, dpi = 300)
```

### Behavioural

```{r}
# Get positions of countries per regime
regime_spans <- behav_pca_score %>%
  arrange(regionWelfare, cntry) %>%
  distinct(cntry, regionWelfare) %>%
  mutate(cntry_id = as.numeric(factor(cntry, levels = unique(cntry)))) %>%
  group_by(regionWelfare) %>%
  summarise(x_start = min(cntry_id), x_end = max(cntry_id), .groups = "drop")

# Get positions of the lines for regime averages 
regime_spans_behavWR <- regime_spans %>% 
  left_join(behav_pca_pooledWR, by = "regionWelfare")


#plot
ggplot(behav_pca_score %>% 
         arrange(regionWelfare) %>% 
         mutate(cntry = factor(cntry, levels = unique(cntry))),
       aes(x = cntry, y = pc1_behav, ymin = ci_l, ymax = ci_u, color = sex)) + 
  geom_pointrange(size = 0.28,
                  aes(alpha = ifelse(sex == "All", 1, 0.4)) ) +
  scale_color_manual(
    values = c("All" = "black", "Men" = "#1F77B4", "Women" = "#E69F00"),
    breaks = c("All", "Men", "Women")
  ) +
  scale_alpha_identity() + 
  geom_hline(yintercept = behav_pca_pooledCountry, linetype = "dotted", color = "black") +
  geom_segment(
    data = regime_spans_behavWR,
    aes(x = x_start, xend = x_end, y = wr_mean_behav, yend = wr_mean_behav),
    inherit.aes = F, linetype = "11", linewidth = 0.7, color = "blue")  +
  geom_segment( 
    data = regime_spans,
    aes(x = x_start, xend = x_end, y = -1.55, yend = -1.55),
    inherit.aes = FALSE,
    size = 1.2,
    color = "grey40"
  ) +
  coord_cartesian(ylim = c(-1.5, 1.5)) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Behavioural disadvantage", 
    y = "Composite Behavioural Score (PCA)"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.x = element_blank(),
        plot.background   = element_rect(fill = "white", color = NA),
        panel.background  = element_rect(fill = "white", color = NA),
        legend.position = "none")

ggsave("Figures\\CompositeScores\\BehavWR_psp.png", width = 6, height = 4, dpi = 300)
```

### Psychosocial

```{r}
# Get positions of countries per regime
regime_spans <- psych_pca_score %>%
  arrange(regionWelfare, cntry) %>%
  distinct(cntry, regionWelfare) %>%
  mutate(cntry_id = as.numeric(factor(cntry, levels = unique(cntry)))) %>%
  group_by(regionWelfare) %>%
  summarise(x_start = min(cntry_id), x_end = max(cntry_id), .groups = "drop")

# Get positions of the lines for regime averages 
regime_spans_psychWR <- regime_spans %>% 
  left_join(psych_pca_pooledWR, by = "regionWelfare")

#plot
ggplot(psych_pca_score %>% 
         arrange(regionWelfare) %>% 
         mutate(cntry = factor(cntry, levels = unique(cntry))),
       aes(x = cntry, y = pc1_psych, ymin = ci_l, ymax = ci_u, color = sex)) + 
  geom_pointrange(size = 0.28,
                  aes(alpha = ifelse(sex == "All", 1, 0.4)) ) +
  scale_color_manual(
    values = c("All" = "black", "Men" = "#1F77B4", "Women" = "#E69F00"),
    breaks = c("All", "Men", "Women")
  ) +
  scale_alpha_identity() +  
  geom_hline(yintercept = psych_pca_pooledCountry, linetype = "dotted", color = "black") +
  geom_segment(
    data = regime_spans_psychWR,
    aes(x = x_start, xend = x_end, y = wr_mean_psych, yend = wr_mean_psych),
    inherit.aes = F, linetype = "11", linewidth = 0.7, color = "blue")  +
  geom_segment( 
    data = regime_spans,
    aes(x = x_start, xend = x_end, y = -1.55, yend = -1.55), 
    inherit.aes = FALSE,
    size = 1.2,
    color = "grey40"
  ) +
  coord_cartesian(ylim = c(-1.5, 1.5)) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Psychosocial disadvantage", 
    y = "Composite Psychosocial Score (PCA)"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.x = element_blank(),
        plot.background   = element_rect(fill = "white", color = NA),
        panel.background  = element_rect(fill = "white", color = NA),
        legend.position = "none")

ggsave("Figures\\CompositeScores\\psychWR_psp.png", width = 6, height = 4, dpi = 300)
```

### Social Capital
```{r}
# Get positions of countries per regime
regime_spans <- sc_pca_score %>%
  arrange(regionWelfare, cntry) %>%
  distinct(cntry, regionWelfare) %>%
  mutate(cntry_id = as.numeric(factor(cntry, levels = unique(cntry)))) %>%
  group_by(regionWelfare) %>%
  summarise(x_start = min(cntry_id), x_end = max(cntry_id), .groups = "drop")

# Get positions of the lines for regime averages 
regime_spans_scWR <- regime_spans %>% 
  left_join(sc_pca_pooledWR, by = "regionWelfare")

#plot

ggplot(sc_pca_score %>% 
         arrange(regionWelfare) %>% 
         mutate(cntry = factor(cntry, levels = unique(cntry))),
       aes(x = cntry, y = pc1_sc, ymin = ci_l, ymax = ci_u, color = sex)) + 
  geom_pointrange(size = 0.28,
                  aes(alpha = ifelse(sex == "All", 1, 0.4)) ) +
  scale_color_manual(
    values = c("All" = "black", "Men" = "#1F77B4", "Women" = "#E69F00"),
    breaks = c("All", "Men", "Women")
  ) +
  scale_alpha_identity() +  
  geom_hline(yintercept = sc_pca_pooledCountry, linetype = "dotted", color = "black") +
  geom_segment(
    data = regime_spans_scWR,
    aes(x = x_start, xend = x_end, y = wr_mean_sc, yend = wr_mean_sc),
    inherit.aes = F, linetype = "11", linewidth = 0.7, color = "blue")  +
  geom_segment( 
    data = regime_spans,
    aes(x = x_start, xend = x_end, y = -1.55, yend = -1.55), 
    inherit.aes = FALSE,
    size = 1.2,
    color = "grey40"
  ) +
  coord_cartesian(ylim = c(-1.7, 1.7)) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Social Capital disadvantage",
    y = "Composite Social Capital Score (PCA)"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.x = element_blank(),
        plot.background   = element_rect(fill = "white", color = NA),
        panel.background  = element_rect(fill = "white", color = NA),
        legend.position = "none")

ggsave("Figures\\CompositeScores\\SCWR_psp.png", width = 6, height = 4, dpi = 300)
```

## SPIDER-PLOTS: PCA

### Prepare data

```{r}

# Select relevant variables

pca_spider_vars <- c("index_edu", "pc1_work", "pc1_mat", "pc1_hc", "pc1_behav", "pc1_psych" , "pc1_sc") 

# Rename variables for the graph

PCAblock_labels <- c(
  index_edu = "Education",
  pc1_work = "Employment", 
  pc1_mat = "Material",
  pc1_hc = "HC access",
  pc1_behav = "Behaviour", 
  pc1_psych = "Psychosocial",
  pc1_sc = "Social Capital"
)


### PCA data European average: prepare ----

all_pca_score_EuMean <- c( 
  index_edu   = edu_index_pooledCountry,
  pc1_work  = work_pca_pooledCountry,
  pc1_mat   = mat_pca_pooledCountry,
  pc1_hc    = hc_pca_pooledCountry,
  pc1_behav = behav_pca_pooledCountry,
  pc1_psych = psych_pca_pooledCountry,
  pc1_sc    = sc_pca_pooledCountry
)


### PCA data WR average ----

# merge allWR_pca_score into a new dataframe 

allWR_pca_score <- merge(edu_index_pooledWR, work_pca_pooledWR, by = c("regionWelfare"))
allWR_pca_score <- merge(allWR_pca_score, mat_pca_pooledWR, by = c("regionWelfare"))
allWR_pca_score <- merge(allWR_pca_score, hc_pca_pooledWR, by = c("regionWelfare"))
allWR_pca_score <- merge(allWR_pca_score, behav_pca_pooledWR, by = c("regionWelfare"))
allWR_pca_score <- merge(allWR_pca_score, psych_pca_pooledWR, by = c("regionWelfare"))
allWR_pca_score <- merge(allWR_pca_score, sc_pca_pooledWR, by = c("regionWelfare"))


#convert all_pca_score to numeric 

allWR_pca_numeric <- allWR_pca_score %>% 
  dplyr::select(c(wr_mean_edu, wr_mean_work, wr_mean_mat, wr_mean_hc, wr_mean_behav, wr_mean_psych, wr_mean_sc )) %>% 
  as.data.frame() %>% 
  lapply(function(x) {
    if (inherits(x, "haven_labelled")) {
      as.numeric(as_factor(x))  
    } else {
      as.numeric(x)
    }
  }) %>% 
  as.data.frame()


### PCA data Country: prepare ----

# clean _pca_score dataframes to only leave cntry and pca score

edu_index_tomerge <- edu_index_pooledsex %>% 
  dplyr::select(cntry, index_edu)
work_pca_tomerge <- work_pca_pooledsex %>% 
  dplyr::select(cntry, pc1_work)
mat_pca_tomerge <- mat_pca_pooledsex %>% 
  dplyr::select(cntry, pc1_mat)
hc_pca_tomerge <- hc_pca_pooledsex %>% 
  dplyr::select(cntry, pc1_hc)
behav_pca_tomerge <- behav_pca_pooledsex %>% 
  dplyr::select(cntry, pc1_behav)
psych_pca_tomerge <- psych_pca_pooledsex %>% 
  dplyr::select(cntry, pc1_psych)
sc_pca_tomerge <- sc_pca_pooledsex %>% 
  dplyr::select(cntry, pc1_sc)

# merge all _pca_score into a new dataframe 

all_pca_score <- merge(edu_index_tomerge, work_pca_tomerge, by = c("cntry"))
all_pca_score <- merge(all_pca_score, mat_pca_tomerge, by = c("cntry"))
all_pca_score <- merge(all_pca_score, hc_pca_tomerge, by = c("cntry"))
all_pca_score <- merge(all_pca_score, behav_pca_tomerge, by = c("cntry"))
all_pca_score <- merge(all_pca_score, psych_pca_tomerge, by = c("cntry"))
all_pca_score <- merge(all_pca_score, sc_pca_tomerge, by = c("cntry"))


#convert all_pca_score to numeric 

all_pca_numeric <- all_pca_score %>% 
  dplyr::select(all_of(pca_spider_vars)) %>% 
  as.data.frame() %>% 
  lapply(function(x) {
    if (inherits(x, "haven_labelled")) {
      as.numeric(as_factor(x))  
    } else {
      as.numeric(x)
    }
  }) %>% 
  as.data.frame()


### Max/Min for each variable ----
# Compute per-axis anchors (max and min) following order

pca_max_val <- sapply(all_pca_numeric[pca_spider_vars], max, na.rm = TRUE)
pca_min_val <- sapply(all_pca_numeric[pca_spider_vars], min, na.rm = TRUE)
```

### Country plots

```{r}
# Loop
for (country in country_list) {
  
  # Filter country-specific data
  country_data <- all_pca_score %>% 
    filter(cntry == country) %>% 
    dplyr::select(all_of(pca_spider_vars))%>% 
    as.data.frame()
  
  # enforce the same column order explicitly 
  country_data <- country_data[, pca_spider_vars, drop = FALSE] 
  
  
  # construct radar df
  radarPCA_df <- rbind(
    pca_max_val[pca_spider_vars],  
    pca_min_val[pca_spider_vars], 
    country_data,    
    all_pca_score_EuMean[pca_spider_vars]  
  )
  
  
  colnames(radarPCA_df) <-  PCAblock_labels[pca_spider_vars] 
  par(lend = 2)   
  
  # Showing plot 
  par(mar = c(1, 2, 3, 2))  
  
  
  radarchart(radarPCA_df,
             axistype = 1,
             
             pcol = c("darkred", "darkblue"),
             pfcol = c(scales::alpha("darkred", 0.2), NA), 
             plwd = c(2, 2),
             cglcol = "grey", 
             cglty = 1,
             axislabcol = "black", 
             #caxislabels = rep("", length(caxislabels_seq)), 
             cglwd = 0.8,
             vlcex = 0.9,
             title = country)
  
  Sys.sleep(1)  
  
  # Saving plot to PNG
  png(filename = paste0("Figures\\pca_per_country_spider\\Spiderplot_", country, ".png"), width = 600, height = 600)
  par(mar = c(1, 2, 3, 2))  
  radarchart(radarPCA_df,
             axistype = 1,
             pcol = c("darkred", "darkblue"), 
             pfcol = c(scales::alpha("darkred", 0.2), NA),
             plwd = c(4, 4),
             cglcol = "grey", 
             cglty = 1,
             axislabcol = "black", 
             #caxislabels = rep("", length(caxislabels_seq)),
             cglwd = 0.8,
             vlcex = 0.9,
             title = country,
             cex.main = 4) 
  dev.off()
  
}

```

### EU plots

```{r}

# construct radar df
radarPCA_empty_df <- as.data.frame(rbind(
  pca_max_val[pca_spider_vars],  ##put max/min of the whole
  pca_min_val[pca_spider_vars], 
  all_pca_score_EuMean[pca_spider_vars])  
)

colnames(radarPCA_empty_df) <-  PCAblock_labels[pca_spider_vars] 

# Show plot 
par(lend = 2)   
par(mar = c(1, 2, 2, 2))  


radarchart(radarPCA_empty_df,
           axistype = 1,
           pcol = "darkblue",
           plwd = 2,
           cglcol = "grey", 
           cglty = 1,
           axislabcol = "black", 
           #caxislabels = rep("", length(caxislabels_seq)), 
           cglwd = 0.8,
           vlcex = 1)

# Saving plot to PNG
png(filename = paste0("Figures\\pca_per_country_spider\\Spiderplot_Empty.png"), width = 600, height = 600)
par(mar = c(1, 2, 2, 2))  
radarchart(radarPCA_empty_df,
           axistype = 1,
           pcol = "darkblue", 
           plwd = 4,
           cglcol = "grey", 
           cglty = 1,
           axislabcol = "black", 
           #caxislabels = rep("", length(caxislabels_seq)), 
           cglwd = 0.8,
           vlcex = 1.8,
           cex.main = 4)
dev.off()
```

### WR plot

```{r}

#Loop

for (WR in WR_list) {
  
  # Filter country-specific data
  WR_data <- allWR_pca_score %>% 
    filter(regionWelfare == WR) %>% 
    dplyr::select(wr_mean_edu, wr_mean_work, wr_mean_mat, wr_mean_hc, wr_mean_behav, wr_mean_psych, wr_mean_sc)%>% 
    as.data.frame()
  
  # construct radar df
  radarPCA_WR_df <- rbind(
    pca_max_val[pca_spider_vars], 
    pca_min_val[pca_spider_vars], 
    WR_data,
    all_pca_score_EuMean[pca_spider_vars] 
  )
  
  colnames(radarPCA_WR_df) <-  PCAblock_labels[pca_spider_vars] 
  
  
  # Show plot 
  par(lend = 2)   
  par(mar = c(1, 2, 3, 2))
  
  
  radarchart(radarPCA_WR_df,
             axistype = 1,
             pcol = c("blue", "darkblue"),
             pfcol = c(scales::alpha("blue", 0.2), NA), 
             plwd = c(2, 2),
             cglcol = "grey", 
             cglty = 1,
             axislabcol = "black", 
             #caxislabels = rep("", length(caxislabels_seq)), 
             cglwd = 0.8,
             vlcex = 0.9,
             title = WR_labels[as.character(WR)])
  Sys.sleep(1)
  
  # Saving plot to PNG
  png(filename = paste0("Figures\\pca_per_country_spider\\Spiderplot_", WR, ".png"), width = 600, height = 600)
  par(mar = c(1, 2, 3, 2))
  radarchart(radarPCA_WR_df,
             axistype = 1,
             pcol = c("blue", "darkblue"),
             pfcol = c(scales::alpha("blue", 0.2), NA), 
             plwd = c(4, 4),
             cglcol = "grey",
             cglty = 1,
             axislabcol = "black",
             #caxislabels = rep("", length(caxislabels_seq)),
             cglwd = 0.8,
             vlcex = 0.9,
             title = WR_labels[as.character(WR)],
             cex.main = 4)
  dev.off()
  
}
```


## CLUSTER ANALYSIS
```{r}
# Dataframe to use
all_pca_score_std <- all_pca_score

# Set country names as rownames
rownames(all_pca_score_std) <- all_pca_score_std$cntry

# Keep only PCA variables and standardize
all_pca_score_std <- scale(all_pca_score_std[, 2:8])

# Distance matrix
all_pca_distance <- dist(all_pca_score_std)

# Hierarchical clustering
clust_all_pca <-hclust(all_pca_distance, method = "ward.D2")

# Dendrogram
countries_wrongWR <- c("CY", "GR", "PL", "SI", "EE", "CH", "NL")

clust_all_pca_dendro <- as.dendrogram(clust_all_pca)
labels_colors(clust_all_pca_dendro) <- ifelse(labels(clust_all_pca_dendro) %in% countries_wrongWR, "red", "black")
plot(clust_all_pca_dendro,
     xlab = "", ylab = "", sub = "")

png("Figures\\Cluster\\Cluster_Ward_D2.png", width = 800, height = 600)
plot(clust_all_pca_dendro,
     xlab = "", ylab = "", sub = "")
dev.off()
```

## ANOVA

```{r}
# Create dataset (from Spiderplot section)
all_pca_score_WR <- all_pca_score %>%
  left_join(country_regionWelfare, by = "cntry")
all_pca_score_WR$regionWelfare <- as.factor(all_pca_score_WR$regionWelfare)
```

### Education

```{r}
aov_edu <- aov(index_edu ~ regionWelfare, data = all_pca_score_WR)

shapiro.test(residuals(aov_edu)) 
leveneTest(index_edu ~ regionWelfare, data = all_pca_score_WR) 

summary(aov_edu) 
```

### Employment Conditions

```{r}
aov_work <- aov(pc1_work ~ regionWelfare, data = all_pca_score_WR) 

shapiro.test(residuals(aov_work)) 
leveneTest(pc1_work ~ regionWelfare, data = all_pca_score_WR) 

summary(aov_work) 
```


### Material

```{r}
aov_mat <- aov(pc1_mat ~ regionWelfare, data = all_pca_score_WR) 

shapiro.test(residuals(aov_mat)) 
leveneTest(pc1_mat ~ regionWelfare, data = all_pca_score_WR)

summary(aov_mat) 
```

### Healthcare

```{r}
aov_hc <- aov(pc1_hc ~ regionWelfare, data = all_pca_score_WR) 

shapiro.test(residuals(aov_hc)) 
leveneTest(pc1_hc ~ regionWelfare, data = all_pca_score_WR) 

summary(aov_hc) 
```

### Behavioural

```{r}
aov_behav <- aov(pc1_behav ~ regionWelfare, data = all_pca_score_WR) 

shapiro.test(residuals(aov_behav)) 
leveneTest(pc1_behav ~ regionWelfare, data = all_pca_score_WR) 

aov_behav <- kruskal.test(pc1_behav ~ regionWelfare, 
                         data = all_pca_score_WR)
summary(aov_behav) 
```

### Psychosocial

```{r}
aov_psych <-aov(pc1_psych ~ regionWelfare, data = all_pca_score_WR) 

shapiro.test(residuals(aov_psych)) 
leveneTest(pc1_psych ~ regionWelfare, data = all_pca_score_WR) 

summary(aov_psych) 
```

### Social Capital
```{r}
aov_sc <- aov(pc1_sc ~ regionWelfare, data = all_pca_score_WR) 

shapiro.test(residuals(aov_sc)) 
leveneTest(pc1_sc ~ regionWelfare, data = all_pca_score_WR) 

summary(aov_sc) 
```
